In-silico molecular enrichment and clearance of the human intracranial space

### 2025-06-10 paper/main.tex
More work on Methods

% Limitations
In terms of limitations, our primary constraint modelling-wise is the
geometric representation of the vascular (and therefore also
perivascular) networks. Concretely, we account for no vessels below
the original MRI resolution limits, and thus only a part of the
cerebral surface vasculature and effectively not the cortical
vessels. In addition, and again due to lack of resolution, there is
uncertainty associated with the location of the vessel segments
relative to the CSF spaces and parenchyma, with some vessels
surrounded by both domains and some vessels completely entering the
parenchyma before returning to the SAS. Higher resolution MRI data
(T1w and e.g. time-of-flight, QSM or other modalities) in more
subjects would immediately allow for more detailed studies of this
aspect. As a result of the lack of an accurate representation of the
cortical and subcortical white matter vasculature, we have considered
a conservative transport model within the parenchyma, modelling
extracellular diffusion alone. To compensate, we have focused on
reporting detailed predictions in the CSF spaces and surface PVS after
not more than 24 hours. In particular, we note that, we observe tracer
accumulation at the perivascular network ends after 12 hours,
similar to that previously observed for
microspheres~\cite{bedussi2018paravascular, mestre2018flow}, which
here can be interpreted as an artifact of the incomplete network
creating a barrier for tracer movement. Previously, we have found
that enrichment within the parenchyma is well-represented by $\approx
3\times$ enhanced diffusion augmented by local clearance e.g.~across
the blood-brain barrier or diffusion-convection with a complex pattern
of spatially-varying convective velocities of 1-10 \textmu
m/min~\cite{vinje2023human}. We will incorporate such transport
mechanisms within the parenchyma in future work. Finally, we also note
that we simulate the intracranial distribution of a tracer
concentration appearing at the craniocervical junction rather than as
an intrathecal injection. We therefore overestimate the amount of
tracer entering into the intracranial spaces; however, due to the
linearity of the mathematical model, it is valid to directly interpret
the in-silico concentrations relative to the total amount of tracer.

@@

The pulsatile flow of CSF in the SAS and ventricular system,
associated with the cardiac and respiratory cycles, substantially
enhances molecular diffusion through
dispersion~\cite{taylor1953dispersion, watson1983diffusion,
  asgari2016glymphatic, sharp2019dispersion, ray2021quantitative,
  troyetsky2021dispersion}. To account for these dispersive effects,
while bridging from the pulsatile flow time scale of seconds to a
molecular transport time scale of hours, we estimate cardiac- and
respiratory dispersion factors ($R_c$ and $R_r$) in $\Omega_{\rm CSF}$
(\Cref{fig:csf}G, I). To compute these spatially-varying factors, we
combine computational estimates of peak CSF pressure gradients with
theoretical estimates for shear-augmented (Taylor) dispersion
\cite{taylor1953dispersion, watson1983diffusion, sharp2019dispersion}
(see Supplementary information S1.3 Estimating dispersion factors from
pulsatile CSF flow for a complete description). Combining the cardiac
and respiratory effects, we set $R = R_c + R_r$ in $\Omega_{\rm CSF}$.

@@

Using the theoretical framework introduced by Gjerde et
al.~\cite{gjerde2023directional}, we also compute analytic estimates
for the time-average perivascular flow rate $\langle Q' \rangle$
induced by peristalsis in the arterial network $\Lambda_a$; i.e.~the
net flow induced by traveling waves of arterial wall motion of
frequency $f$, amplitude $\varepsilon$ and wave lengths $\lambda$
(\Cref{fig:pvs_a}C). The corresponding contribution to the
periarterial flow velocity is defined for each segment $\Lambda_i$ as
$\langle Q'_i \rangle/A_i$ where $A_i = \pi (R_2^i - R_1^i)$ is the
cross-section area of the PVS segment and $\langle Q_i' \rangle$ is
the estimated mean flow rate (see Supplementary information S1.4 for
more details). Two wall motion patterns are considered, yielding
$\hat{u}_a^c$ and $\hat{u}_a^v$ respectively: \emph{cardiac}
pulsations with $f = 1.0$Hz, $\lambda = 2$ m and $\varepsilon = 1\%$
(\Cref{fig:pvs_a}D), and very low-frequency \emph{vasomotion} with $f
= 0.1$Hz, $\lambda = 0.02$m, and $\varepsilon = 10\%$
(\Cref{fig:pvs_a}E). No flow induced by peristalsis is considered for
the perivenous network. In the baseline model, we combine the
contributions from CSF production and cardiac peristalsis to set
$\hat{u}$ in~\eqref{eq:multi_transport} as $\hat{u}_a = \hat{u}_a^p +
\hat{u}_a^c$ in $\Lambda_a$, and $\hat{u}_v = \hat{u}_v^p$ in
$\Lambda_v$. We remark that none of these models account for fluid
exchange across the lateral outer interface of the PVSs (over
$\Gamma_a, \Gamma_v$), while fluid entry and exit is allowed at the
end nodes of the PVS networks.

>>>>

% Limitations
In terms of limitations, our primary constraint modelling-wise is the
geometric representation of the vascular (and therefore also
perivascular) networks. Concretely, we account for no vessels below
the original MRI resolution limits, and thus only a part of the
cerebral surface vasculature and effectively not the cortical
vessels. In addition, and again due to lack of resolution, there is
uncertainty associated with the location of the vessel segments
relative to the CSF spaces and parenchyma, with some vessels
surrounded by both domains and some vessels completely entering the
parenchyma before returning to the SAS. Higher resolution MRI data
(T1w and e.g. time-of-flight, QSM or other modalities) in more
subjects would immediately allow for more detailed studies of this
aspect. As a result of the lack of an accurate representation of the
cortical and subcortical white matter vasculature, we have considered
a basic transport model within the parenchyma, modelling extracellular
diffusion alone. To compensate, we have focused on reporting detailed
predictions in the CSF spaces and surface PVS after not more than 24
hours. In particular, we note that, we observe tracer accumulation at
the perivascular network ends after 12 hours, similar to that
previously observed for microspheres~\cite{bedussi2018paravascular,
  mestre2018flow}, which here can be interpreted as an artifact of the
incomplete network creating a barrier for tracer movement. Previously,
by inferring transport parameters within the human brain from
glymphatic MRI data, we have found that molecular transport within the
parenchyma is well-represented by $\approx 3\times$ enhanced diffusion
augmented by local clearance e.g.~across the blood-brain barrier, or
by diffusion-convection with a complex pattern of spatially-varying
interstitial fluid flow on the order of 1-10 \textmu
m/min~\cite{vinje2023human}. We will incorporate such transport
mechanisms within the parenchyma in future work. Finally, we also note
that we simulate the intracranial distribution of a tracer
concentration appearing at the craniocervical junction rather than as
an intrathecal injection. We therefore overestimate the amount of
tracer entering into the intracranial spaces; however, due to the
linearity of the mathematical model, it is valid to directly interpret
the in-silico concentrations relative to the total amount of tracer.

@@

The pulsatile flow of CSF in the SAS and ventricular system,
associated with the cardiac and respiratory cycles, substantially
enhances molecular diffusion through
dispersion~\cite{taylor1953dispersion, watson1983diffusion,
  asgari2016glymphatic, sharp2019dispersion, ray2021quantitative,
  troyetsky2021dispersion}. To account for these dispersive effects,
while bridging from the pulsatile flow time scale of seconds to a
molecular transport time scale of hours, we estimate cardiac- and
respiratory dispersion factors ($R_c$ and $R_r$) in $\Omega_{\rm CSF}$
(\Cref{fig:csf}G, I). To compute these spatially-varying factors, we
combine computational estimates of peak CSF pressure gradients
(\Cref{fig:csf}F, H) with theoretical estimates for shear-augmented
(Taylor) dispersion \cite{taylor1953dispersion, watson1983diffusion,
  sharp2019dispersion} (see Supplementary information S1.3 \emph{Estimating
dispersion factors from pulsatile CSF flow} for a complete
description). Combining the cardiac and respiratory effects, we set $R
= R_c + R_r$ in $\Omega_{\rm CSF}$.

@@

Using the theoretical framework introduced by Gjerde et
al.~\cite{gjerde2023directional}, we also compute analytic estimates
for the time-average perivascular flow rate $\langle Q' \rangle$
induced by peristalsis in the arterial network $\Lambda_a$; i.e.~the
net flow induced by traveling waves of arterial wall motion of
frequency $f$, amplitude $\varepsilon$ and wave lengths $\lambda$
(\Cref{fig:pvs_a}C). The corresponding contribution to the
periarterial flow velocity is defined for each segment $\Lambda_i$ as
$\langle Q'_i \rangle/A_i$ where $A_i = \pi (R_2^i - R_1^i)$ is the
cross-section area of the PVS segment and $\langle Q_i' \rangle$ is
the estimated mean flow rate (see Supplementary information S1.4
\emph{Estimating net perivascular flow induced by peristaltic waves}
for more details). Two wall motion patterns are considered:
\emph{cardiac} pulsations with $f = 1.0$Hz, $\lambda = 2$ m and
$\varepsilon = 1\%$ yielding $\hat{u}_a^c$ (\Cref{fig:pvs_a}D), and
very low-frequency \emph{vasomotion} with $f = 0.1$Hz, $\lambda =
0.02$m, and $\varepsilon = 10\%$ yielding $\hat{u}_a^v$
(\Cref{fig:pvs_a}E). No flow induced by peristalsis is considered for
the perivenous network. In the baseline model, we combine the
contributions from CSF production and cardiac peristalsis to set
$\hat{u}$ as $\hat{u}_a = \hat{u}_a^p + \hat{u}_a^c$ in $\Lambda_a$,
and $\hat{u}_v = \hat{u}_v^p$ in $\Lambda_v$. We remark that none of
these models account for fluid exchange across the lateral outer
interface of the PVSs (over $\Gamma_a, \Gamma_v$), while fluid entry
and exit is allowed at the end nodes of the PVS networks.
